Sparse Function-space Representation \\ of Neural Networks

### 2023-05-15 paper/main.tex
added rl to related work

The contributions of this paper are:
%
{\em (i)}~We introduce \our, a new approach for sparse functional representation of the neural network.
{\em (ii)}~We demonstrate that, despite its sparsity, our method effectively captures predictive uncertainty, provides means of updating the model post-training, and gives a a compact representation suitable for continual learning.
{\em (iii)}~We provide extensive experiments for showcasing our approach and demonstrate significance and applicability across supervised, continual, and reinforcement learning, aiming to stimulate future use of the approach.

@@

@@

\textbf{Linearization gives rise to a Gaussian process}
Our goal is to capture the distribution over the neural network model functions through their first two moments. The first moments characterize a Gaussian process with a mean function $\mu(\cdot)$ and a covariance function (kernel) $\kappa(\cdot,\cdot)$. 
Recent work \cite{khan2019approximate,maddox2021fast} has shown that linearizing approximations in the weight space lead to function-space equivalent approximations.
As Gaussian distributions remain tractable under linear transformations, a linear function in terms of parameters can be converted from the weight space to the function space (see Ch.~2.1 in \cite{rasmussen2006gaussian}) as follows:
%
\begin{equation} \label{eq:weight_func}
f_\weights(\vx) \approx 
%g_\weights(\mathbf{x}) = 
\phi^\top\!(\vx) \, \vw \quad\implies\quad \mu(\vx) = 0 \quad \text{and} \quad \kappa(\vx, \vx') = \frac{1}{\delta} \phi^\T\!(\vx) \, \phi(\vx').
\end{equation}
% The posterior structure directly relates to the optimization loss around the MAP weights $\vw^*$.
A common approach is to approximate the correlation structure of a distribution centred at the MAP estimate as done in the Laplace-GGN~\cite{khan2019approximate, daxberger2021laplace, maddox2021fast}. 
The Laplace-GGN takes the MAP solution and approximates the Hessian of the loss function
with the GGN.
Following the work of \citet{khan2019approximate}, we can view this approximation as building an approximate linear model of the neural network as $f_{\weights^*}(\vx) \approx 
%g_{\weights}(\vx) = 
\Jac{\weights_*}{\vx} \, \weights$, where $\Jac{\weights}{\vx} \coloneqq \left[ \nabla_\weights f_\weights(\vx)\right]^\top \in \R^{C \times P}$ is the Jacobian at the MAP.
Using the Hessian of the approximate model, we arrive at the Laplace-GGN approximate posterior for $q(\vw)$. 
Therefore, the Laplace-GGN linear model can be used to convert our weight space model to the function space by,
\begin{equation} 
\label{eq-laplace-approx-function-space} 
% g(\vx) \sim \GP \left( \mu(\vx), \kappa(\vx, \vx') \right) \quad \text{with} \quad
  \mu(\vx) =  0 \quad \text{and} \quad
  \kappa(\vx, \vx')
  = \frac{1}{\delta} \Jac{\weights^*}{\vx} \, \JacT{\weights^*}{\vx'}, 
\end{equation}
where the kernel is the so-called Neural Tangent Kernel (NTK, \cite{jacot2018neural}). In \cref{eq-laplace-approx-function-space}, we restrict the function output to a single dimension for notational simplicity, yet this formulation can be readily extended to the multi-output scenario. With this kernel function, we can combine it with the data set $\dataset$ to construct the posterior $q(\vf)$. 
%
\citet{khan2019approximate} took a similar approach, but instead of fitting the GP posterior to the actual $\vy$, they rely on a transformation of $\vy$ referred as {\em pseudo}-data. Similarly, \citet{immer2021improving} obtain the same covariance function but use a different mean because they rely on a first-order approximation of the neural network to form a Bayesian GLM model. Both approaches attempted to adjust the GP posterior mean function to ensure the predictions are from the neural network $f_{\vw^*}(\vx)$. 
Contrarily, our methodology intends to make predictions directly from the derived GP, thus evading such adjustments. Next, we will present the general formulation of the sparse functional representation of our trained neural network based on the non-sparse approximation.

@@

As was shown in \cite{adam2021dual}, the posterior under this model has a particular structure similar to \cref{eq:gp_pred}. In the paper, the authors spot the dual variables' importance for approximate inference, but write them with the natural parameterization as this form is more suitable for optimization through natural gradients. In order to link the dual variables defined in \cref{eq:dual_param}, we write them in the sparse GP using the dual variables
\begin{equation} 
	\valpha_{\vu}  =  \sum_{i \in \mathcal{D}_{n}}  \vkzi \, \hat{\alpha}_{i}, \quad
	\vbeta_{\vu} =  \sum_{i \in \mathcal{D}_{n}} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    
\label{eq:dual_sparse}
\end{equation}
where the sparse dual variables are now a sum over all data points, with $\valpha_{\vu} \in \R^{M}$ and $\vbeta_{\vu} \in \R^{M  \times M}$. Using these sparse definition of the dual variables, our sparse GP posterior takes the following form:
\begin{equation}\label{eq:dual_sparse_post}
   \myexpect_{q_{\vu}(\vf)}[f_i] = \vkzs^{\T} \MKzz^{-1}   \valpha_{\vu} , \quad 
   \textrm{Var}_{q_{\vu}(\vf)}[f_i]  = \kappa_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \vbeta_{\vu})^{-1} ]\vkzs
\end{equation}
where $\MKzz$ and $\vkzs$ are defined similarly to $\MKxx$ and $\vk_{\vx i}$ but over the inducing points $\MZ \in \R^{M \times D} $  instead of the full data set $\dataset$. The key quantities we need to make predictions from our sparse GP from the converged neural network are $(\hat{\alpha}_i, \hat{\beta}_i)$ \cref{eq:dual_param_laplac}\todo{these variables need another subscript for differentiate that from the values computed at the data points} and a kernel function $\kappa$ . Contrasting \cref{eq:dual_sparse_post} and \cref{eq:gp_pred}, we can see that the computational complexity went from $\mathcal{O}(n^3)$ to $\mathcal{O}(m^3)$, with $m \ll n$.  Crucially, given the structure of our probabilistic model, our sparse dual variables \cref{eq:dual_sparse} are a compact representation of the full model projected using the kernel.

@@

\subsection{\our for Continual Learning}
% Definition of CL -> a sequence of tasks 
In the Continual learning setting, the training is divided into $T$ tasks, each with its training data set $\dataset_t = \{(\mathbf{x}_{i}, \mathbf{y}_{i})\}_{i=1}^{n_t}$, which after the task is discarded and cannot be accessed in the following stages of the training. Rehearsal and function regularization-based models keep a subset of the trai ning data for each task to help the model alleviate forgetting. In the same way, after each task, we can use our \our to build a compact representation of the neural network and then exploit this representation to build a functional regularizer to add to the training loss for the subsequent tasks. Recently proposed GP-based regularizers \cite{ pan2020continual, rudner2022continual} have shown better performances than weight space equivalents getting state-of-the-art on CL benchmarks. However, they are limited to use subset approximations of the posterior.
%, which in practice means randomly selecting some data computing its posterior at only those points and then using that as the regularizer. 
A more principled approach would be to sparsify the resultant GP into a low-rank approximation to ensure information coming from the entire task dataset is not lost. %, essentially what the sparse dual variables do.   \todo{this term is replacing the missing data}
% Where the regularizer comes from
At the end of the training for each task $t$, we compute the Laplace-GGN approximated kernel function for the current model and randomly select $m$ inducing points. Given the issues described in the previous sections, related to computing the full covariance posterior $\MKxx$ for $\dataset_t$,\todo{this should be $\kappa(\MX_t, \MX_t)$} we can exploit the dual parameterization to efficiently encode the information from  $\dataset_t$ using the dual parameterization, as follows, 
\begin{equation}
 	\bar{\MB}^{-1}_t = \MKzz^{-1} \vbeta_\vu \MKzz^{-1} \in \R^{m \times m}, 
 	\quad 
 	\vbeta_{\vu} =  \sum_{i \in \dataset_t} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    
 	% \quad \forall t \in [1, T]
 \end{equation}
 where $\MKzz$ and $\vkzi$ are the kernel function computed on the inducing points $\MZ_t \in \R^{m \times D}$ for task $t$ .

We keep a memory buffer $\mathcal{M} = \{(\MZ_t, \vu_t, \bar{\MB}^{-1}_t)\}_{t=1}^T$, where $\MZ_t$ are the inducing points selected from $\dataset_t$ and  $\vu_t = f_{\vw_t^*}(\MZ_t)\in \R^{m}$ are the neural network outputs associated to them, computed with the MAP estimate after task $t$ $\weights_t^*$. Note that, for notational convenience, we are extending our neural network notation to be compatible with multiple inputs points, \ie, each $i$th entry is $f_{\weights_t^*}(\vz_i)\in \R$. Moreover, we restrict ourselves to single-output neural networks and refer the reader to further details on the multi-output setting to \cref{sec:cl_multioutput}.

@@

\begin{equation}
	\mathcal{R}_\textit{SFR}(\weights, \mathcal{M}) = \frac{1}{2} \sum_{s=1}^{t-1} \frac{1}{m} 
	\left\lVert 
	f_{\weights}(\MZ_{s}) - \vu_t % f_{\weights_{s}}(\MZ_s)
	\right\rVert_{\bar{\MB}^{-1}_{s}}.
\end{equation}
%\todo{choose one option}
% where $f_\vw(\MZ_{s}) \in \R^m$ are the outputs of neural networks with the current weights $\vw$ and stored function outputs of the neural network at the MAP optimum for task $s$, i.e., each $i$th entry is $f_\cdot(\vz_i) \in \R$.
%\begin{equation}
%	\argmin_{\weights} \sum_{i=1}^{n} \ell(f_\weights(\mathbf{x}_{i}), \mathbf{y}_i) + \delta \mathcal{R}(\weights) + \tau \mathcal{R}_\textit{SFR}(\weights)
%\end{equation}
Then, when the training on the successive task, the new objective to minimize takes the following form
\begin{equation}
	\weights_t^* = \argmin_{\weights} \mathcal{L}(\mathcal{D}_t, \weights) + \tau \mathcal{R}_\textit{SFR}(\weights, \mathcal{M})
\end{equation}

@@

\subsection{Updating the  network representation for Continual learning}
% the problem is hard many hyperparams and also a lot of things can make huge differences
% we restrict ourselves to the SH setting, then some values

>>>>

The contributions of this paper are:
%
{\em (i)}~We introduce \our, a new approach for sparse representation of the neural network in the function space.
{\em (ii)}~We demonstrate that, despite its sparsity, our method effectively captures predictive uncertainty, provides means of updating the model post-training, and gives a a compact representation suitable for continual learning.
{\em (iii)}~We provide extensive experiments for showcasing our approach and demonstrate significance and applicability across supervised, continual, and reinforcement learning, aiming to stimulate future use of the approach.

@@

\textbf{Uncertainty quantification in RL}
A key challenge in RL is balancing the trade-off between exploration and exploitation \cite{sutton2018reinforcement}.
That is, should an agent select actions that it knows will lead to high reward (exploitation), or should it
select new ones in hope to discover actions leading to higher reward (exploration).
One promising direction to balancing this trade-off is to model the uncertainty associated with a learned transition dynamics model and use it to guide exploration.
Prior work has taken an expectation over the dynamic model's posterior \cite{deisenrothPILCO2011,kamtheDataEfficient2018,chuaDeepReinforcementLearning2018},
sampled from it (akin to Thompson sampling but referred to as posterior sampling RL)
\todo{cite thopmson sampling paper dearden et al 1999}
\cite{osbandMoreEfficientReinforcement2013},
and used it to implement optimism in the face of uncertainty using upper confidence bounds \cite{curiEfficient2020}.
\todo{cite UCRL paper}
No single strategy has emerged as a go to method and we highlight that these strategies are only as good as their uncertainty estimates.
% Central to these decision-making strategies are the methods for quantifying uncertainty.
Previous approaches have used GPs \cite{deisenrothPILCO2011,kamtheDataEfficient2018},
ensembles of neural networks \cite{curiEfficient2020,chuaDeepReinforcementLearning2018}
and variational inference \cite{galImproving2016,houthooftVIME2017}.
However, there has been little leveraging

% Another classical and provable exploration theory in RL is based on optimism
% \todo{cite UCRL paper}
% Recent work has even extended these ideas to deep RL \cite{curiEfficient2020}.

% An alternative approach is to sample from the dynamic model's posterior, akin to Thompson sampling but referred to as posterior sampling RL
% Another classical and provable exploration theory in RL is based on optimism in the face of uncertainty.
% \todo{cite UCRL paper}
% Recent work has even extended these ideas to deep RL \cite{curiEfficient2020}.
% These strategies are only as good as their uncertainty estimates.
% Prior work has learned dynamics models using GPs \cite{deisenrothPILCO2011,kamtheDataEfficient2018},
% ensembles of neural networks \cite{curiEfficient2020,chuaDeepReinforcementLearning2018}
% and variational inference \cite{galImproving2016,houthooftVIME2017}.

@@

\textbf{Linearization gives rise to a Gaussian process}
We are concerned with capturing the distribution over the neural network model functions, in terms of the first two moments of the distribution. A Gaussian process is fully characterized by these moments: a mean function $\mu(\cdot)$ and a covariance (kernel) function $\kappa(\cdot,\cdot)$. 
Recent work \cite{khan2019approximate,maddox2021fast} has shown that linearizing approximations in the weight space lead to function-space equivalent approximations.
As Gaussian distributions remain tractable under linear transformations, a linear function in terms of parameters can be converted from the weight space to the function space (see Ch.~2.1 in \cite{rasmussen2006gaussian}) as follows:\begin{equation} \label{eq:weight_func}
f_\weights(\vx) \approx 
%g_\weights(\mathbf{x}) = 
\phi^\top\!(\vx) \, \vw \implies \mu(\vx) = 0 \quad \kappa(\vx, \vx') = \frac{1}{\delta} \phi^\T\!(\vx) \, \phi(\vx')
\end{equation}
% The posterior structure directly relates to the optimization loss around the MAP weights $\vw^*$.
A common approach is to approximate the correlation structure of a distribution centred at the MAP estimate as done in the Laplace-GGN \citep{khan2019approximate, daxberger2021laplace, maddox2021fast}. The Laplace-GGN takes the MAP solution and approximates the Hessian of the loss function
% $\ell(f_\weights(\mathbf{x}_{i}), \mathbf{y}_i)$
with the GGN.
Following the work of \citep{khan2019approximate}, we can view this approximation as building an approximate linear model of the neural network as $f_{\weights^*}(\vx) \approx 
%g_{\weights}(\vx) = 
\Jac{\weights_*}{\vx} \, \weights$, where $\Jac{\weights}{\vx} \coloneqq \left[ \nabla_\weights f_\weights(\vx)\right]^\top \in \R^{C \times P}$ is the Jacobian at the MAP.
Using the Hessian of the approximate model, we arrive at the Laplace-GGN approximate posterior for $q(\vw)$.
\todo{do we need this sentence about the Hessian to get Laplace-GGN??}
Therefore, the Laplace-GGN linear model can be used to convert our weight space model to the function space by,
\begin{equation} 
\label{eq-laplace-approx-function-space} 
% g(\vx) \sim \GP \left( \mu(\vx), \kappa(\vx, \vx') \right) \quad \text{with} \quad
  \mu(\vx) =  0 \quad \text{and} \quad
  \kappa(\vx, \vx')
  = \frac{1}{\delta} \Jac{\weights^*}{\vx} \, \JacT{\weights^*}{\vx'}, 
\end{equation}
where the kernel is the so-called Neural Tangent Kernel (NTK, \cite{jacot2018neural}). In \cref{eq-laplace-approx-function-space}, for notational convenience, we restricted the function output to single dimension, but this could be easily extended for the multi-dimensional case. With this kernel function, we can then combine it with the dataset $\dataset$ to form the posterior $q(\vf)$. \citet{khan2019approximate} took a similar approach, but instead of fitting the GP posterior to the actual $\vy$, they rely on a transformation of $\vy$ referred as pseudo-data. Similarly, \citet{immer2021improving} obtain the same covariance function but a different mean because they rely on a first order approximation of the neural network to form a Bayesian GLM model. Both approaches attempted to adjust the GP posterior mean function to ensure the predictions are from the neural network $f_{\vw^*}(\vx)$. In our method, we wish to make predictions from the derived GP and thus we avoid such adjustments. We now present the general formulation of the sparse functional representation of our trained neural network based on the non-sparse approximation.

@@

As was shown in \cite{adam2021dual}, the posterior under this model has a particular structure similar to \cref{eq:gp_pred}. In the paper, the authors spot the dual variables' importance for approximate inference, but write them with the natural parameterization as this form is more suitable for optimization through natural gradients. In order to link the dual variables defined in \cref{eq:dual_param}, we write them in the sparse GP using the dual variables
\begin{equation} 
	\valpha_{\vu}  =  \sum_{i \in \mathcal{D}_{n}}  \vkzi \, \hat{\alpha}_{i}, \quad
	\vbeta_{\vu} =  \sum_{i \in \mathcal{D}_{n}} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    
\label{eq:dual_sparse}
\end{equation}
where the sparse dual variables are now a sum over all data points, with $\valpha_{\vu} \in \R^{M}$ and $\vbeta_{\vu} \in \R^{M  \times M}$. Using these sparse definition of the dual variables, our sparse GP posterior takes the following form:
\begin{equation}\label{eq:dual_sparse_post}
   \myexpect_{q_{\vu}(\vf)}[f_i] = \vkzs^{\T} \MKzz^{-1}   \valpha_{\vu} , \quad 
   \textrm{Var}_{q_{\vu}(\vf)}[f_i]  = \kappa_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \vbeta_{\vu})^{-1} ]\vkzs
\end{equation}
where $\MKzz$ and $\vkzs$ are defined similarly to $\MKxx$ and $\vk_{\vx i}$ but over the inducing points $\MZ \in \R^{M \times D} $  instead of the full dataset $\dataset$. The key quantities we need to make predictions from our sparse GP from the converged neural network are $(\hat{\alpha}_i, \hat{\beta}_i)$ \cref{eq:dual_param_laplac}\todo{these variables need another subscript for differentiate that from the values computed at the data points} and a kernel function $\kappa$ . Contrasting \cref{eq:dual_sparse_post} and \cref{eq:gp_pred}, we can see that the computational complexity went from $\mathcal{O}(n^3)$ to $\mathcal{O}(m^3)$, with $m \ll n$.  Crucially, given the structure of our probabilistic model, our sparse dual variables \cref{eq:dual_sparse} are a compact representation of the full model projected using the kernel.

@@

\subsection{\our for Continual Learning}
% Definition of CL -> a sequence of tasks 
In the Continual learning setting, the training is divided into $T$ tasks, each with its training data set $\dataset_t = \{(\mathbf{x}_{i}, \mathbf{y}_{i})\}_{i=1}^{n_t}$, which after the task is discarded and cannot be accessed in the following stages of the training. Rehearsal and function regularization-based models keep a subset of the training data for each task to help the model alleviate forgetting. In the same way, after each task, we can use our \our to build a compact representation of the neural network and then exploit this representation to build a functional regularizer to add to the training loss for the subsequent tasks. Recently proposed GP-based regularizers \cite{ pan2020continual, rudner2022continual} have shown better performances than weight space equivalents getting state-of-the-art on CL benchmarks. However, they are limited to use subset approximations of the posterior.
%, which in practice means randomly selecting some data computing its posterior at only those points and then using that as the regularizer. 
A more principled approach would be to sparsify the resultant GP into a low-rank approximation to ensure information coming from the entire task dataset is not lost.%, essentially what the sparse dual variables do.   \todo{this term is replacing the missing data}
% Where the regularizer comes from
At the end of the training for each task $t$, we compute the Laplace-GGN approximated kernel function for the current model and randomly select $m$ inducing points. Given the issues described in the previous sections, related to computing the full covariance posterior $\MKxx$,\todo{this should be $\kappa(\MX_t, \MX_t)$}, we can exploit the dual parameterization to efficiently encode the information from the $\dataset_t$ using the dual parameterization, as follows, 
\begin{equation}
 	\bar{\MB}^{-1}_t = \MKzz^{-1} \vbeta_\vu \MKzz^{-1} \in \R^{m \times m}, 
 	\quad 
 	\vbeta_{\vu} =  \sum_{i \in \dataset_t} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    
 	% \quad \forall t \in [1, T]
 \end{equation}
 where $\MKzz$ and $\vkzi$ are the kernel function computed on the inducing points $\MZ_t \in \R^{m \times D}$ for task $t$.

We can then keep a memory buffer $\mathcal{M} = \{(\MZ_t, \vu_t, \bar{\MB}^{-1}_t)\}_{t=1}^T$, where $\MZ_t$ are the inducing points selected from $\dataset_t$ and  $\vu_t = f_{\vw_t^*}(\MZ_t)\in \R^{m}$ are the neural network outputs over them computed with the MAP estimate after task $t$, $\vw_t^*$. Note that, for notational convenience, we are extending our neural network notation to be compatible with multiple inputs points, \ie, each $i$th entry is $f_{\vw_t^*}(\vz_i)\in \R$. Moreover, we restrict ourselves to single-output neural networks and refer the reader to further details on the multi-output setting to \cref{sec:cl_multioutput}.

@@

\begin{equation}
	\mathcal{R}_\textit{SFR}(\weights, \mathcal{M}) = \sum_{s=1}^{t-1} \frac{1}{m} 
	\left\lVert 
	f_{\weights}(\MZ_{s}) - \vu_t % f_{\weights_{s}}(\MZ_s)
	\right\rVert_{\bar{\MB}^{-1}_{s}}.
\end{equation}
%\todo{choose one option}
% where $f_\vw(\MZ_{s}) \in \R^m$ are the outputs of neural networks with the current weights $\vw$ and stored function outputs of the neural network at the MAP optimum for task $s$, i.e., each $i$th entry is $f_\cdot(\vz_i) \in \R$.
%\begin{equation}
%	\argmin_{\weights} \sum_{i=1}^{n} \ell(f_\weights(\mathbf{x}_{i}), \mathbf{y}_i) + \delta \mathcal{R}(\weights) + \tau \mathcal{R}_\textit{SFR}(\weights)
%\end{equation}
Then, when the training on the successive task, the new objective to minimize takes the following form
\begin{equation}
	\weights_t^* = \argmin_{\weights} \mathcal{L}(\mathcal{D}_t, \weights) + \tau \mathcal{R}_\textit{SFR}(\weights, \mathcal{M})
\end{equation}

@@

\subsection{Updating the  network representation for Continual learning}
The Continual learning setting is a challenging and the training and
